---
title: Introduction to R and the Tidyverse
author: Clemens Schmid
format:
  html:
    link-external-icon: true
    link-external-newwindow: true
editor_options: 
  chunk_output_type: console
bibliography: assets/references/chapters/r-tidyverse/references.bib
---

::: callout-note
This session is typically ran held in parallel to the Introduction to Python and Pandas. Participants of the summer schools chose which to attend based on their prior experience. We recommend this session if you have no experience with neither R nor Python.
:::

::: callout-tip
For this chapter's exercises, if not already performed, you will need to create the [conda environment](before-you-start.qmd#creating-a-conda-environment) from the `yml` file in the following [link](https://github.com/SPAAM-community/intro-to-ancient-metagenomics-book/raw/main/assets/envs/r-tidyverse.yml) (use `wget`, or right click and save as to download), and once created, activate the environment with:

``` bash
conda activate r-tidyverse
```

To download the data for this chapter, please download following archive with, extract the tar, and change into the directory.

For example

```bash
wget https://www.dropbox.com/s/61zy5uadsatxz7y/r-tidyverse.tar.gz -P /<PATH>/<TO>/<SOMEWHERE_TO_STORE>/
tar -xzf r-tidyverse.tar.gz
cd r-tidyverse
```
:::

```{r, echo=FALSE}
# Set global options
knitr::opts_chunk$set(attr.output = "style='border: 1px; border-style: solid; margin-left: 10px; margin-right: 10px;'")
```

## The working environment

### R, RStudio, the tidyverse and penguins

R [@RCoreTeam2023] is a fully featured programming language, but it excels as an environment for (statistical) data analysis (<https://www.r-project.org>)
RStudio [@RstudioTeam] is an integrated development environment (IDE) for R (and other languages) (<https://www.rstudio.com/products/rstudio>)

The tidyverse [@Wickham2019-ot] is a powerful collection of R packages with well-designed and consistent interfaces for the main steps of data analysis: loading, transforming and plotting data (<https://www.tidyverse.org>). This tutorial works with tidyverse ~v2.0. We will learn about the packages `readr`, `tibble`, `ggplot2`, `dplyr`, `magrittr` and `tidyr`. `forcats` will be briefly mentioned, but `purrr` and `stringr` are left out. 

The `palmerpenguins` package [@Horst2020] provides a neat example dataset to learn data exploration and visualization in R (<https://allisonhorst.github.io/palmerpenguins>)

::: {.callout-note title="Instructions for running this chapter without conda" collapse=true}

If you wish to follow this chapter without installing conda, you can recreate the conda environment and data generation by running the following steps.

1. Install R and RStudio for your operating system according to one of the many instructions online (e.g. [here](https://www.r-bloggers.com/2022/01/how-to-install-and-update-r-and-rstudio))
2. Download the .qmd file underlying this webpage [here](https://github.com/SPAAM-community/intro-to-ancient-metagenomics-book/blob/main/r-tidyverse.qmd) and copy it to a new directory
3. Open RStudio and create a new project in this directory with `File` -> `New Project...` -> `Existing directory`
4. Open this very file from the RStudio project, so that you can easily follow along
5. Run the `Environment preparation code` code to install necessary R package dependencies and prepare the required data used below


```{r, eval=FALSE}
# installing necessary R packages
install.packages(c("tidyverse", "palmerpenguins"))

# preparing data
# (at the end of this tutorial
# you will understand this code)
library(magrittr)
set.seed(5678)

peng_prepped <- palmerpenguins::penguins %>%
    dplyr::filter(
        !dplyr::if_any(
            .cols = tidyselect::everything(),
            .fns = is.na
        )
    ) %>%
    tibble::add_column(., id = 1:nrow(.), .before = "species")

peng_prepped %>%
    dplyr::slice_sample(n = 300) %>%
    dplyr::arrange(id) %>%
    dplyr::select(-bill_length_mm, -bill_depth_mm) %>%
    readr::write_csv("penguins.csv")

peng_prepped %>%
    dplyr::slice_sample(n = 300) %>%
    dplyr::arrange(id) %>%
    dplyr::select(id, bill_length_mm, bill_depth_mm) %>%
    readr::write_csv("penguin_bills.csv")
```

:::

To load RStudio from within your `conda` environment, please run

```bash
rstudio
```

And the RStudio window will open.

## Loading data into tibbles

### Reading data with readr

With R we usually operate on data in our computer's memory.
The tidyverse provides the package `readr` to read data from text files into the memory.
`readr` can read from our file system or the internet.
It provides functions to read data in almost any (text) format:

```{r eval=FALSE}
readr::read_csv() # .csv files -> see penguins.csv
readr::read_tsv() # .tsv files
readr::read_delim() # tabular files with an arbitrary separator
readr::read_fwf() # fixed width files
readr::read_lines() # read linewise to parse yourself
```

### How does the interface of `read_csv` work?

We can learn more about an R function with the `?` operator.

To open a help file for a specific function run `?<function_name>` (e.g. `?readr::read_csv`) in the R console.
`readr::read_csv` has many options to specify how to read a text file.

```{r eval=FALSE}
read_csv(
  file,                      # The path to the file we want to read
  col_names = TRUE,          # Are there column names?
  col_types = NULL,          # Which types do the columns have? NULL -> auto
  locale = default_locale(), # How is information encoded in this file?
  na = c("", "NA"),          # Which values mean "no data"
  trim_ws = TRUE,            # Should superfluous white-spaces be removed?
  skip = 0,                  # Skip X lines at the beginning of the file
  n_max = Inf,               # Only read X lines
  skip_empty_rows = TRUE,    # Should empty lines be ignored? 
  comment = "",              # Should comment lines be ignored?
  name_repair = "unique",    # How should "broken" column names be fixed
  ...
)
```

### What does `readr` produce? The `tibble`!

To read a .csv file (here `"penguins.csv"`) into a variable (here `peng_auto`) run.

```{r, eval=FALSE}
peng_auto <- readr::read_csv("penguins.csv")
```

```{r, echo=FALSE}
# this version is only for the website!
peng_auto <- readr::read_csv("assets/data/r-tidyverse/penguins.csv")
```

`readr` first prints some information on the number and type of rows and columns it discovered in the file.

It automatically detects column types -- but you can also define them manually.

```{r, eval=FALSE}
peng <- readr::read_csv(
  "penguins.csv",
  col_types = "iccddcc" # this string encodes the desired types for each column
)
```

```{r, echo=FALSE}
# this version is only for the website!
peng <- readr::read_csv(
  "assets/data/r-tidyverse/penguins.csv",
  col_types = "iccddcc" # this string encodes the desired types for each column
)
```

It then returns an in-memory representation of this data, a `tibble`.

A `tibble` is a "data frame", a tabular data structure with rows and columns.
Unlike a simple array, each column can have another data type.

### How to look at a `tibble`?

Typing the name of any object into the R console will print an overview to the console.

```{r}
peng
```

But there are various other ways to inspect the content of a `tibble` 

```{r, eval=FALSE}
str(peng) # a structural overview of an R object
summary(peng) # a human-readable summary of an R object
View(peng) # open RStudio's interactive data browser
```

## Plotting data in `tibble`s

### `ggplot2` and the "grammar of graphics"

To understand and present data, we usually have to visualize it.

`ggplot2` is an R package that offers an unusual, but powerful and logical interface for this task [@Wickham2016]. The following example describes a stacked bar chart.

```{r}
library(ggplot2) # Loading a library to use its functions without ::
```

```{r}
ggplot( # Every plot starts with a call to the ggplot() function
    data = peng # This function can also take the input tibble
) + # The plot consists of individual functions linked with "+"
    geom_bar( # "geoms" define the plot layers we want to draw
        mapping = aes( # The aes() function maps variables to visual properties
            x = island, # publication_year -> x-axis
            fill = species # community_type   -> fill color
        )
    )
```

A `geom_*` combines data (here `peng`), a geometry (here vertical, stacked bars) and a statistical transformation (here counting the number of penguins per island and species). `ggplot2` features many such geoms: A good overview is provided by this cheatsheet: [https://rstudio.github.io/cheatsheets/html/data-visualization.html](https://rstudio.github.io/cheatsheets/html/data-visualization.html).

Beyond `geom`s, a ggplot plot can be further specified with (among others) `scale`s, `facet`s and `theme`s.

### `scale`s control the behaviour of visual elements

Here is another plot to demonstrate this: Boxplots of penguin weight per species.

```{r}
ggplot(peng) +
    geom_boxplot(aes(x = species, y = body_mass_g))
```

Let's assume we had some extreme outliers in this dataset. To simulate this, we replace some random weights with extreme values.

```{r}
set.seed(1234) # we set a seed for reproducible randomness
peng_out <- peng
peng_out$body_mass_g[sample(1:nrow(peng_out), 10)] <- 50000 + 50000 * runif(10)
```

Now we plot the dataset with these "outliers".

```{r}
ggplot(peng_out) +
    geom_boxplot(aes(x = species, y = body_mass_g))
```

This is not well readable, because the extreme outliers dictate the scale of the y-axis. A 50+kg penguin is a scary thought and we would probably remove these outliers, but let's assume they are valid observation we want to include in the plot.

To mitigate this issue we can change the **scale** of different visual elements - e.g. the y-axis.

```{r}
ggplot(peng_out) +
    geom_boxplot(aes(x = species, y = body_mass_g)) +
    scale_y_log10() # adding the log-scale improves readability
```

We will now go back to the normal dataset without the artificial outliers.

### Colour `scale`s

(Fill) colour is a visual element of a plot and its scaling can be adjusted as well.

```{r}
ggplot(peng) +
    geom_boxplot(aes(x = species, y = body_mass_g, fill = species)) +
    scale_fill_viridis_d(option = "C")
```

We use the `scale_*` function to select one of the visually appealing (and robust to colourblindness) viridis colour palettes ([https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html](https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html)).

### More variables! Defining plot matrices via `facet`s

In the previous example we didn't add additional information with the fill colour, as the plot already distinguished by species on the x-axis.

We can instead use colour to encode more information, for example by mapping the variable sex to it.

```{r}
ggplot(peng) +
    geom_boxplot(aes(x = species, y = body_mass_g, fill = sex))
```

Another way to visualize more variables in one plot is to split the plot by categories into **facets**, so sub-plots per category.

Here we split by sex, which is already mapped to the fill colour.

```{r}
ggplot(peng) +
    geom_boxplot(aes(x = species, y = body_mass_g, fill = sex)) +
    facet_wrap(~sex)
```

The fill colour is therefore free again to show yet another variable, for example the year a given penguin was examined.

```{r}
ggplot(peng) +
    geom_boxplot(aes(x = species, y = body_mass_g, fill = year)) +
    facet_wrap(~sex)
```

### Setting purely aesthetic settings with `theme`

Aesthetic changes can be applied as part of the `theme`, which allows for very detailed configuration (see `?theme`).

Here we rotate the x-axis labels by 45°, which often helps to resolve overplotting.

```{r}
ggplot(peng) +
    geom_boxplot(aes(x = species, y = body_mass_g, fill = year)) +
    facet_wrap(~sex) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```

### Ordering elements in a plot with `factors`

R supports defining ordinal data with `factor`s. This can be used to set the order of elements in a plot, e.g. the order of bars in a bar chart. 

We do not cover `factor`s beyond the following example here, although the tidyverse includes a package (`forcats`) specifically for that purpose.

Elements based on `character` columns are generally ordered alphabetically.

```{r}
ggplot(peng) +
    geom_bar(aes(x = species)) # bars are alphabetically ordered
```

With `forcats::fct_reorder` we can transform an input vector to a `factor`, ordered by a summary statistic (even based on another vector).

```{r}
peng2 <- peng
peng2$species_ordered <- forcats::fct_reorder(
  peng2$species,
  peng2$species, length
)
```

With this change, the plot will be ordered according to the intrinsic order defined for `species_ordered`.

```{r}
ggplot(peng2) +
    geom_bar(aes(x = species_ordered)) # bars are ordered by size
```

### Exercise

1. Look at the `mtcars` dataset and read up on the meaning of its variables with the help operator `?`. `mtcars` is a test dataset integrated in R and can be accessed by typing `mtcars` in the console

2. Visualize the relationship between *Gross horsepower* and *1/4 mile time*

```{r}

```

3. Integrate the *Number of cylinders* into your plot as an additional variable

```{r}

```

::: {.callout-tip title="Possible solutions" collapse=true}


```{r, eval=FALSE}
?mtcars
```

```{r}
ggplot(mtcars) +
    geom_point(aes(x = hp, y = qsec))
```

```{r}
ggplot(mtcars) +
    geom_point(aes(x = hp, y = qsec, color = as.factor(cyl)))
```

:::

## Conditional queries on tibbles

### Selecting columns and filtering rows with `select` and `filter`

```{r, echo=FALSE}
# technical adjustments for rendering
old_options <- options(
  pillar.print_max = 5,
  pillar.print_min = 5,
  pillar.advice = FALSE
)
```

The `dplyr` package includes powerful functions to subset data in tibbles based on conditions. `dplyr::select` allows to select columns.

```{r}
dplyr::select(peng, id, flipper_length_mm) # reduce tibble to two columns
dplyr::select(peng, -island, -flipper_length_mm) # remove two columns
```

`dplyr::filter` allows for conditional filtering of rows.

```{r}
dplyr::filter(peng, year == 2007) # penguins examined in 2007
dplyr::filter(peng, year == 2007 |
    year == 2009) # penguins examined in 2007 OR 2009
dplyr::filter(
    peng, # an alternative way to express
    year %in% c(2007, 2009)
) # OR with the match operator "%in%"
dplyr::filter(peng, species == "Adelie" &
    body_mass_g >= 4000) # Adelie penguins heavier than 4kg
```

### Chaining functions together with the pipe `%>%`

A core feature of the tidyverse is the pipe `%>%` in the `magrittr` package. This 'infix' operator allows to chain data and operations for concise and clear data analysis syntax.

```{r}
library(magrittr)
peng %>% dplyr::filter(year == 2007)
```

It forwards the LHS (left-hand side) as the first argument of the function appearing on the RHS (right-hand side), which enables sequences of functions ("tidyverse style").

```{r}
peng %>%
    dplyr::select(id, species, body_mass_g) %>%
    dplyr::filter(species == "Adelie" & body_mass_g >= 4000) %>%
    nrow() # count the resulting rows
```

`magrittr` also offers some more operators, among which the extraction `%$%` is particularly useful to easily extract individual variables.

```{r}
peng %>%
    dplyr::filter(island == "Biscoe") %$%
    species %>% # extract the species column as a vector
    unique() # get the unique elements of said vector
```

Here we already use the base R summary function `unique`.

### Summary statistics in `base` R

Summarising and counting data is indispensable and R offers all basic operations you would expect in its `base` package.

```{r}
chinstraps_weights <- peng %>%
    dplyr::filter(species == "Chinstrap") %$%
    body_mass_g

length(chinstraps_weights) # length/size of a vector
unique(chinstraps_weights) # unique elements of a vector

min(chinstraps_weights) # minimum
max(chinstraps_weights) # maximum

mean(chinstraps_weights) # mean
median(chinstraps_weights) # median

var(chinstraps_weights) # variance
sd(chinstraps_weights) # standard deviation
quantile(chinstraps_weights, probs = 0.75) # quantiles for the given probabilities
```

Many of these functions can ignore missing values with an option `na.rm = TRUE`.

### Group-wise summaries with `group_by` and `summarise`

These summary statistics are particular useful when applied to conditional subsets of a dataset.

`dplyr` allows such summary operations with a combination of the functions `group_by` and `summarise`, where the former tags a `tibble` with categories based on its variables and the latter reduces it to these groups while simultanously creating new columns.

```{r}
peng %>%
    dplyr::group_by(species) %>% # group the tibble by the material column
    dplyr::summarise(
        min_weight = min(body_mass_g), # new col: min weight for each group
        median_weight = median(body_mass_g), # new col: median weight for each group
        max_weight = max(body_mass_g) # new col: max weight for each group
    )
```

Grouping can also be applied across multiple columns at once.

```{r}
peng %>%
    dplyr::group_by(species, year) %>% # group by species and year
    dplyr::summarise(
        n = dplyr::n(), # a new column: number of penguins for each group
        .groups = "drop" # drop the grouping after this summary operation
    )
```

If we group by more than one variable, then `summarise` will not entirely remove the group tagging when generating the result dataset. We can force this with `.groups = "drop"` to avoid undesired behaviour with this dataset later on.

### Sorting and slicing tibbles with `arrange` and `slice`

`dplyr` allows to `arrange` tibbles by one or multiple columns.

```{r}
peng %>% dplyr::arrange(sex) # sort by sex
peng %>% dplyr::arrange(sex, body_mass_g) # sort by sex and weight
peng %>% dplyr::arrange(dplyr::desc(body_mass_g)) # sort descending
```

Sorting also works within groups and can be paired with `slice` to extract extreme values per group.

Here we extract the heaviest individuals per species.

```{r}
peng %>%
    dplyr::group_by(species) %>% # group by species
    dplyr::arrange(dplyr::desc(body_mass_g)) %>% # sort by weight within (!) groups
    dplyr::slice_head(n = 3) %>% # keep the first three penguins per group
    dplyr::ungroup() # remove the still lingering grouping
```

Slicing is also the relevant operation to take random samples from the observations in a `tibble`.

```{r}
peng %>% dplyr::slice_sample(n = 10)
```

### Exercise

For this exercise we once more go back to the `mtcars` dataset. See `?mtcars` for details

1. Determine the number of cars with four *forward gears* (`gear`) in the `mtcars` dataset

```{r}

```

2. Determine the mean *1/4 mile time* (`qsec`) per *Number of cylinders* (`cyl`) group

```{r}

```

3. Identify the least efficient cars for both *transmission types* (`am`)

```{r}

```

::: {.callout-tip title="Possible solutions" collapse=true}

```{r}
mtcars %>%
    dplyr::filter(gear == 4) %>%
    nrow()
```

```{r}
mtcars %>%
    dplyr::group_by(cyl) %>%
    dplyr::summarise(
        qsec_mean = mean(qsec)
    )
```

```{r}
mtcars2 <- tibble::rownames_to_column(mtcars, var = "car")
mtcars2 %>%
    dplyr::group_by(am) %>%
    dplyr::arrange(mpg) %>%
    dplyr::slice_head() %$%
    car
```

:::

## Transforming and manipulating tibbles

### Renaming and reordering columns with `rename` and `relocate`

Columns in tibbles can be renamed with `dplyr::rename`.

```{r}
peng %>% dplyr::rename(penguin_name = id) # rename a column
```

And with `dplyr::relocate` they can be reordered.

```{r}
peng %>% dplyr::relocate(year, .before = species) # reorder columns
```

Adding columns to tibbles with `mutate` and `transmute`.

A common application of data manipulation is adding new, derived columns, that combine or modify the information in the already available columns. `dplyr` offers this core feature with the `mutate` function.

```{r}
peng %>%
    dplyr::mutate( # add a column that
        kg = body_mass_g / 1000 # manipulates an existing column
    )
```

`dplyr::transmute` has the same interface as `dplyr::mutate`, but it removes all columns except for the newly created ones.

```{r}
peng %>%
  dplyr::transmute(
    id = paste("Penguin Nr.", id), # overwrite this column
    flipper_length_mm               # select this column
)
```

`tibble::add_column` behaves as `dplyr::mutate`, but gives more control over column position.

```{r}
peng %>% tibble::add_column(
  flipper_length_cm = .$flipper_length_mm/10, # not the . representing the LHS of the pipe
  .after = "flipper_length_mm"
)
```

`dplyr::mutate` can also be combined with `dplyr::group_by` (instead of `dplyr::summarise`) to add information on a group level. This is relevant, when a value for an individual entity should be put into context of a group-wise summary statistic.

```{r}
peng %>%
    dplyr::group_by(species, sex, year) %>%
    dplyr::mutate(
        mean_weight = mean(body_mass_g, na.rm = T),
        relation_to_mean = body_mass_g / mean_weight
    ) %>%
    dplyr::ungroup() %>%
    dplyr::select(id, species, sex, year, relation_to_mean) %>%
    # mutate does not remove rows, unlike summarise, so we use select
    dplyr::arrange(dplyr::desc(relation_to_mean))
```

### Conditional operations with `ifelse`, `case_when` and `case_match`

`ifelse` allows to implement conditional `mutate` operations, that consider information from other columns.

```{r}
peng %>% dplyr::mutate(
  weight = ifelse(
    test = body_mass_g >= 4200, # is weight below or above mean weight?
    yes = "above mean",
    no = "below mean"
  )
)
```

`ifelse` gets cumbersome easily. `dplyr::case_when` is more readable and scales much better for this application.

```{r}
peng %>% dplyr::mutate(
  weight = dplyr::case_when(
    body_mass_g >= 4200 ~ "above mean", # the number of conditions is arbitrary
    body_mass_g < 4200  ~ "below mean",
    TRUE                ~ "unknown"     # TRUE catches all remaining cases
  )
)
```

`dplyr::case_match` is similar, but unlike `dplyr::case_when` it does not check logical expressions, but matches by value.

```{r}
peng %>%
    dplyr::mutate(
        island_rating = dplyr::case_match(
            island,
            "Torgersen" ~ "My favourite island",
            "Biscoe" ~ "Overrated tourist trap",
            "Dream" ~ "Lost my wallet there. 4/10"
        )
    ) %>%
    # here we use group_by+summarise only to show the result
    dplyr::group_by(island, island_rating) %>%
    dplyr::summarise(.groups = "drop")
```

### Switching between long and wide data with `pivot_longer` and `pivot_wider`

To simplify certain analysis or plotting operations data often has to be transformed from a **wide** to a **long** format or vice versa (@fig-rtidyverse-longtowide).

![Graphical representation of converting a table from a wide format (first column are categories, e.g. species, first row are also categories e.g., samples, and counts are in cells after first row/column. Conversion from wide to long makes the long-format so that the first columnn still has categories, but the second column becomes the column-categories from the first row, and third column has values. Converting long format back to wider shows again first row and column as categories and cells as values).](assets/images/chapters/r-tidyverse/pivot_longer_wider.png){#fig-rtidyverse-longtowide height=150px}

- A table in **wide** format has N key columns and N value columns.
- A table in **long** format has N key columns, one descriptor column and one value column.

Here is an example of a wide dataset. It features information about the number of cars sold per year per brand at a dealership.

```{r}
carsales <- tibble::tribble(
    ~brand, ~`2014`, ~`2015`, ~`2016`, ~`2017`,
    "BMW", 20, 25, 30, 45,
    "VW", 67, 40, 120, 55
)
carsales
```

Wide format data becomes a problem, when the columns are semantically identical. This dataset is in wide format and we can not easily plot it. We generally prefer data in long format, although it is more verbose with more duplication.

To transform this dataset to a long format, we can apply `tidyr::pivot_longer`.

```{r}
carsales_long <- carsales %>% tidyr::pivot_longer(
    cols = tidyselect::num_range(
        "",
        range = 2014:2017
    ), # define a set of columns to transform
    names_to = "year", # the name of the descriptor column we want
    names_transform = as.integer, # a transformation function to apply to the names
    values_to = "sales" # the name of the value column we want
)
carsales_long
```

Wide datasets are not always the wrong choice, for example for adjacency matrices to represent graphs, covariance matrices or other pairwise statistics.
When data gets big, then wide formats can be significantly more efficient (e.g. for spatial data).

So transform data from long to wide, we can use `tidyr::pivot_wider`

```{r}
carsales_wide <- carsales_long %>% tidyr::pivot_wider(
    id_cols = "brand", # the set of id columns that should not be changed
    names_from = year, # the descriptor column with the names of the new columns
    values_from = sales # the value column from which the values should be extracted
)
carsales_wide
```

### Exercise

1. Move the column `gear` to the first position of the mtcars dataset

```{r}

```

2. Make a new dataset `mtcars2` with the column `mpg` and an additional column `am_v`, which encodes the *transmission type* (`am`) as either `"manual"` or `"automatic"`

```{r}

```

3. Count the number of cars per *transmission type* (`am_v`) and *number of gears* (`gear`). Then transform the result to a wide format, with one column per *transmission type*.

```{r}

```

::: {.callout-tip title="Possible solutions" collapse=true}

```{r}
mtcars %>%
    dplyr::relocate(gear, .before = mpg) %>%
    tibble::as_tibble() # transforming the raw dataset for better printing
```

```{r}
mtcars2 <- mtcars %>%
    dplyr::mutate(
        gear,
        am_v = dplyr::case_match(
            am,
            0 ~ "automatic",
            1 ~ "manual"
        )
    ) %>%
    tibble::as_tibble()
mtcars2
```

```{r}
mtcars2 %>%
    dplyr::group_by(am_v, gear) %>%
    dplyr::tally() %>% # dplyr::tally() is identical to
    # dplyr::summarise(n = dplyr::n())
    # it counts the number of entities in a group
    tidyr::pivot_wider(
        names_from = am_v,
        values_from = n
    )
```

:::

## Combining tibbles with join operations

### Types of joins

Joins combine two datasets x and y based on overlapping key columns.

Mutating joins add columns from one dataset to the other:

  - Left join: Take observations from x and add fitting information from y.
  - Right join: Take observations from y and add fitting information from x.
  - Inner join: Join the overlapping observations from x and y.
  - Full join: Join all observations from x and y, even if information is missing.

Filtering joins remove observations from x based on their presence in y.

Types of filtering consist of: 

  - Semi join: Keep every observation in x that is in y.
  - Anti join: Keep every observation in x that is not in y.

The following sections will introduce each join with an example.

To experiment with joins, we need a second dataset with complementary information. This new dataset contains additional variables for a subset of the penguins in our first dataset -- both datasets feature 300 penguins, but only with a partial overlap in individuals.

```{r eval=FALSE}
bills <- readr::read_csv("penguin_bills.csv")
```

```{r echo=FALSE,message=FALSE}
# this version is only for the website!
bills <- readr::read_csv("assets/data/r-tidyverse/penguin_bills.csv")
bills
```

### Left join with `left_join`

Take observations from x and add fitting information from y (@fig-rtidyverse-leftjoin).

![Graphical representation of left join operation. Two tables with a shared first column (A B C, and A B D respectively) are merged together to have columns A B C D. As A and B have a one to one match of values, this remains the same in the joined table. The B column between the two have a different value on the third row, and thus is lost from the second table, retaining row 3 of the first table. Column D (from the second table) has an empty value on row three, as this row was not in row 3 of the second table that column D was derived from.](assets/images/chapters/r-tidyverse/left_join.png){height=150px #fig-rtidyverse-leftjoin} 

```{r}
dplyr::left_join(
  x = peng,  # 300 observations
  y = bills, # 300 observations
  by = "id"  # the key column by which to join
)
```

Left joins are the most common join operation: Add information from y to the main dataset x.

### Right join with `right_join`

Take observations from y and add fitting information from x (@fig-rtidyverse-rightjoin).

![Graphical representation of right join operation. Two tables with a shared first column (A B C, and A B D respectively) are merged together to have columns A B C D. As A and B have a one to one match of values, this remains the same in the joined table. The B column between the two have a different value on the third row, and thus is lost from the first table, retaining row 3 of the second table. Column C (from the first table) has an empty value on row three, as this row was not in row 3 of the first table that column C was derived from.](assets/images/chapters/r-tidyverse/right_join.png){height=150px #fig-rtidyverse-rightjoin} 

```{r}
dplyr::right_join(
    x = peng, # 300 observations
    y = bills, # 300 observations
    by = "id"
) %>% # we arrange by id to highlight the missing
    dplyr::arrange(id) # observation in the peng dataset
```

Right joins are almost identical to left joins -- only x and y have reversed roles.

### Inner join with `inner_join`

Join the overlapping observations from x and y (@fig-rtidyverse-innerjoin).

![Graphical representation of inner join operation. Two tables with a shared first column (A B C, and A B D respectively) are merged together to have columns A B C D. Only rows from both table that have exact matches on columns A and B are retained. The third rows from both tables that had a different value in column B are lost.](assets/images/chapters/r-tidyverse/inner_join.png){height=150px #fig-rtidyverse-innerjoin} 

```{r}
dplyr::inner_join(
  x = peng,  # 300 observations
  y = bills, # 300 observations
  by = "id"
)
```

Inner joins are a fast and easy way to check, to which degree two dataset overlap.

### Full join with `full_join`

Join all observations from x and y, even if information is missing (@fig-rtidyverse-fulljoin).

![Graphical representation of full join operation. Two tables with a shared first column (A B C, and A B D respectively) are merged together to have columns A B C D. All rows from both tables are retained, even though they do not share the same value on column B on both tables. The missing values for the two third rows (i.e., column C from the second table, and column D from the first table) are are filled with an empty cell indicated with `-`.](assets/images/chapters/r-tidyverse/inner_join.png){height=150px #fig-rtidyverse-fulljoin} 

```{r}
dplyr::full_join(
    x = peng, # 300 observations
    y = bills, # 300 observations
    by = "id"
) %>% dplyr::arrange(id)
```

Full joins allow to preserve every bit of information.

### Semi join with `semi_join`

Keep every observation in x that is in y (@fig-rtidyverse-semijoin).

![Graphical representation of semi join operation. Two tables with a shared first column (A B C, and A B D respectively) are merged together to have only columns A B C. Only columns A B and C are retained in the joined table. Row three of both tables are not included as the column values in columns A and B do not match.](assets/images/chapters/r-tidyverse/semi_join.png){height=150px #fig-rtidyverse-semijoin} 

```{r}
dplyr::semi_join(
  x = peng,  # 300 observations
  y = bills, # 300 observations
  by = "id"
)
```

Semi joins are underused (!) operations to filter datasets.

### Anti join with `anti_join`

Keep every observation in x that is not in y (@fig-rtidyverse-antijoin).

![Graphical representation of anti join operation. Two tables with a shared first column (A B C, and A B D respectively) are merged together to have only columns A B C, and of that only row 3 of the first table. Only row three is retained from the first table as this is the only unique row present only in the first table.](assets/images/chapters/r-tidyverse/anti_join.png){height=150px #fig-rtidyverse-antijoin} 


```{r}
dplyr::anti_join(
  x = peng,  # 300 observations
  y = bills, # 300 observations
  by = "id"
)
```

Anti joins allow to quickly determine what information is missing in a dataset compared to an other one.

### Exercise 

Consider the following additional dataset with my opinions on cars with a specific number of gears:

```{r}
gear_opinions <- tibble::tibble(
  gear = c(3, 5),
  opinion = c("boring", "wow")
)
```

1. Add my opinions about gears to the `mtcars` dataset

```{r}

```

2. Remove all cars from the dataset for which I do not have an opinion

```{r}

```

::: {.callout-tip title="Possible solutions" collapse=true}

```{r}
dplyr::left_join(mtcars, gear_opinions, by = "gear") %>%
    tibble::as_tibble()
```

```{r}
dplyr::anti_join(mtcars, gear_opinions, by = "gear") %>%
    tibble::as_tibble()
```

:::

## (Optional) clean-up

Let's clean up your working directory by removing all the data and output from this chapter.

When closing `rstudio`, say no to saving any additional files.

The command below will remove the `/<PATH>/<TO>/r-tidyverse` directory **as well as all of its contents**. 

:::{.callout-tip}
## Pro Tip
Always be VERY careful when using `rm -r`. Check 3x that the path you are
specifying is exactly what you want to delete and nothing more before pressing
ENTER!
:::

```bash
rm -r /<PATH>/<TO>/r-tidyverse*
```

Once deleted you can move elsewhere (e.g. `cd ~`).

We can also get out of the `conda` environment with

```bash
conda deactivate
```

## References

::: {#refs}
:::